Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru

### 2020-08-10 text/ecography/round_3/reply_to_reviewers_2.Rmd

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}

title: "Point-by-point reply to reviewers' comments"
bibliography: ../transect_refs.bib
csl: ../ecography.csl
# biblio-style: "natbib"
link-citations: true
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
library("RSQLite")
library("sf")
# make sure that figures are allowed to fload (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
# needed for dates in experiment figure species richness, otw months are displayed in German
Sys.setlocale("LC_TIME", "C")
```

# Reply to subject editor Bethany Bradley

Dear Dr Bradley,

thank you very much for handling our manuscript.
We are pleased to resubmit a revised version of our manuscript entitled  *Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru*.
We are also very thankful for the constructive comments of both reviewers which helped to improve substantially our manuscript.
Please find below a point-by-point reply to all concerns raised by the reviewers.
In accordance with your and the reviewers' suggestions, we have:

- integrated into a coherent story the vegetation transect analysis and the irrigation-fertiization experiment.
- deleted parts of the discussion to discuss above all findings supported by our data.
- completely rewritten the Abstract, the Conclusions and in great parts the Introduction to make our aims and the corresponding results clearer.
The Conclusions now exclusively feature the 'take-home' points providing an answer to the hypotheses posed in the introduction. 
- added a new figure showing how species richness developed in the experimental study.

After having addressed thoroughly and carefully the reviewers' comments for the third time, we now hope that you deem our manuscript suitable for publication in *Ecography*.
In any case, we would like to express our gratitude again towards you and the reviewers for taking the time to consider and revise our manuscript.

Yours sincerely,

The authors

# Reply to reviewer #1

1. **The different studies are not well combined and appear too much in parallel, especially in the discussion section. Also it is unclear why the spatial prediction was performed because it does not appear to be needed to discuss the findings. In general, the manuscript needs a lot more clarification in the presentation of the aim, hypotheses, and discussion of the results.**

Thanks for taking the time to revise and improve our manuscript.
We have integrated into a single story our transect analysis and the water-fertilizer experiment.
Please refer to the replies to comment #7, 11, 23.

The spatial prediction is crucial for this study.
We have added a new paragraph to the introduction, described the reasoning in more detail and highlighted the resulting findings more prominently throughout the whole manuscript. 
Please refer also the replies to comments #16-19.

Furthermore, we have rewritten our hypotheses, clarified the general aim of our study, have deleted parts of the discussion and have completely rewritten our Conclusions mirroring now the posed hypotheses, and hence the structure of the manuscript.
Please refer to the replies to comments #7-14, 23, 26.

2. **The abstract is not very clearly written and is not presenting a good summary of the paper. E.g. the first two sentences are a poor motivation for this study (the motivation is to study TDF only because they are understudied?). It would be better to start with a motivation that highlight the relevance, e.g. that these areas feature unique biodiversity but that the impact of ENSO on biodiversity patterns is still unknown.**

Thanks for the advice.
We have rewritten as follows:

>Tropical dry forests (TDF) are heavily shaped by the El Niño Southern Oscillation (ENSO) in NW Peru, where especially El Niño episodes bring rain to arid to semi-arid areas.
However, the resulting effects on biodiversity patterns remain largely unknown as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Therefore, we studied the spatio-temporal effects of different ENSO episodes on floristic biodiversity along a climatic gradient ranging from the coastal desert to the Andean foothills.
We sampled 50 vegetation plots in four years representing different ENSO episodes.
To highlight the spatio-temporal changes in floristic composition and beta diversity across ENSO episodes, we predicted ordination scores with a Generalized Additive Model.
We applied variation partitioning to test if topographic or edaphic variables gained in importance during more humid ENSO episodes.
Additionally, we executed an irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction under different simulated ENSO rainfall scenarios. 
Plant species richness increased under humid conditions during the humid La Niña (2012) and the moderate El Niño episodes (2016), and slightly decreased under the very humid conditions during the coastal El Niño episode (2017).
The spatial prediction revealed that specific vegetation formations became more pronounced with increasing water input, but that a large water surplus led to the disruption of the strict order along the climatic gradient.
Edaphic and topographic variables gained in importance with increased water availability (2012 & 2016), however, this effect was not further amplified under very wet conditions (2017).
The experiment showed that plant cover under Super Niño conditions was three times higher when fertilized.
Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics of TDF in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.

3. **Carefully check for logical reasoning in the abstract. E.g. you write that you wanted to test the effect of other variables when water is no longer the limiting factor. But it was not mentioned before that water is a limiting factor.**

Thanks, please refer to the reply to the previous comment.

4. **Your aims and hypotheses other than just studying biodiversity in the TDF (which is quite a broad aim) is not getting clear. Be more specific here on what you want to find out with your study and what you expect.**

Please refer to the reply to comment #2.

5. **It appears unclear at that point how the results of this study “can help the local population from devastating ENSO effects”.**

We have toned down by rewriting the text as follows:

>Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics of TDF in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.

6. **Line 11-15: As your study area is in Peru you should mention what effect the ENSO phases have for your study area instead of a global description of the phenomenon.**

We have rewritten as follows:

>ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
ENSO consists of alternating episodes.
El Niño represents the warm and La Niña the cold episode of this cycle. 
ENSO neutral refers to periods when neither El Niño nor La Niña episodes are present. 
The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, El Niño episodes bring more rain than usual and La Niña episodes enforces the already dry conditions in the study area [<50 mm of annual precipitation along the coast of NW Peru; Fig. 1D;  rollenbeck_climatic_2015].

7. **In general the aim as well as hypotheses are not getting very clear in the introduction. E.g. Line 40: “To tackle the outlined issues...” Be more specific here what issues you’re talking about. In the sentences above you talk about the effect of different soil properties. But you are not addressing this in your methods. Also it’s not getting clear why you want to study edaphic and topographic properties on floristic composition. How does this link to your major aim to study bidoversity with different ENSO phases? This needs a lot more clarification. Also the motivation why different ENSO phases are studied is not getting clear. Make clear why this is relevant.**

Thank you for pushing us to:

- make our aims and hypotheses clearer.
- clarify why it is important to study biodiversity during different ENSO episodes.

We have added several new sentences to the introduction.

With regard to different ENSO episodes, we added:

>Therefore, one objective of this manuscript is to study the effect of different ENSO episodes, including La Niña episodes, on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.

With regard to the predictive mapping, we added a new paragraph (see also the reply to comment #15) while already mentioning the corresponding aim as follows:

>We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.

With regard to the impact of environmental variables, we added:

>Hence, this study investigates how much the impact of edaphic and topographic variables varies in explaining the floristic composition between dry and humid ENSO years.
Additionally, an irrigation-fertilization experiment should help to quantify in more detail the beneficial effects of the water-nutrient interaction on vegetation diversity and cover.

Additionally, we have rewritten the aims of our manuscript as follows:

>In summary, in this manuscript we study the effects of different ENSO episodes on biodiversity patterns and floristic composition  as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Specifically, we tested the following three hypotheses:
>
(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].
>
(2) Water availability limits plant cover and species richness along the entire climatic gradient in the (semi-)arid study area.
Therefore, we were interested in evaluating the effects of soil characteristics and topography changed in shaping floristic composition and vegetation cover (biomass production) with varying rainfall patterns as caused by different ENSO episodes. 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years, i.e., as soon as water is no longer the limiting factor.
>
(3) To quantify the beneficial effects of the water-nutrient interaction on vegetation cover development and species richness, we executed an irrigation-fertilization experiment simulating different ENSO rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño episode in combination with the fertilizer would be most conducive to plant cover and species richness development.

As now written in more detail in our hypotheses, we also wanted to test if the impact of edaphic and topographic variables change as soon as water is no longer the limiting factor in the study area.
Different ENSO episodes create different rainfall scenarios, and therefore provide an excellent opportunity to test if and how the influence of environmental variables change the floristic composition and biodiversity.

Please also note that we describe in section *2.2.2 Environmental variables* which edaphic variables were sampled in the field.

8. **Line 45/46: Don’t mix up hypothesis and methods here**

Please refer to the reply to the previous comment and to comment #18.

9. **Line 50/52: Why do you expect that topography/soil properties have a higher effect on biodiversity in wetter years? The hypothesis is not getting clear.**

Please refer to the reply to comment #7.

10. **Line 53-57: This is a description of the method and unclear what the corresponding aim is. Better describe the hypothesis/aim here and describe the methods applied to approach this later on.**

Please refer to the reply to comment #7.

11. **In the introduction it is unclear what kind of biodiversity you’re studying (not even mentioned that you focus on vegetation!). But that’s relevant for the understanding/hypotheses. E.g. for trees, I would not expect that biodiversity changes during ENSO cycles. So please state what you’re studying and why you expect changes.**

You are absolutely right.
We have added following sentence to the introduction to make clear what we study:

> Therefore, one aim of this manuscript is to study the effect of different ENSO, including La Niña, episodes on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.

Please refer also to the reply to comment #7.

12. **Section 2.2.1: You often talk about plant development. Be more specific here. Do you mean biomass/coverage/diversity?**

We have clarified as follows:

>In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction on species richness, plant cover and biomass production.

13. **2.2.1: Experimental in addition is great to test effect of water availability in a more systematic way than under real-world conditions. But was it vegetation on the experimental plots from the same ecosystem? Or how can the two surveys be compared?**

We have clarified as follows:

>We established the experiment in the middle of the study area on on a protected dryland forest inside the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. 1).
Hence, the experiment was located in the same ecosystem as the landscape study (see previous paragraph) and more specifically in the middle of the observed precipitation gradient, ranging from 26 mm (Paita; annual median precipitation) to 272 mm (Chulucanas; Fig. 1).

Please refer also to Appendix 2 for a more detailed description of the experiment.

14. **Section 2.2.2: Because the hypotheses/aims are not well described in the introduction, it is not clear now why soil samples were taken.**

Please refer to the reply to comment #7.

15. **Section 2.2.2: Why did you use NDVI from MODIS (250m resolution) when you have much smaller plots? Wouldn’t something like Landsat/Sentinel-2 be more appropriate? Also see comment above: Not clear at this point why these variables were calculated because the aims/hypotheses were not made clear. Also at the beginning of the method chapter a short summary would help gaining understanding of what you did for which purposes before going into detail here.**

We have not used Landsat/Sentinel-2 imagery due to following reasons:

- Sentinel-2 data have only become available globally since March 2017.
For consistency reasons, we believe it is better to use the same sensor for each modeled year instead of mixing data from different sensors for different years.
- MODIS data has a much better temporal resolution than Landsat imagery, i.e., as opposed to MODIS data, Landsat data was not available for the vegetation period in each sampled year whereas there were thress MODIS images available for each vegetation period.
Please note also that there is not one pair of locations that share the same MODIS pixel.

16. **2.3.2: Here you mention for the first time, that one aim is to do spatio-temporal prediction. This should have been already mentioned in the introduction. Also there please make clear WHY this is an aim of your study. Predictive mapping even occurs in the title but it is not made clear why you’re doing this, other than  for visualization as you mention in the introduction.**

You are absolutely right, we should have already emphasized in the introduction the importance of predictive mapping, and why we are doing it.
Therefore, we have added following paragraph to the introduction:

>Statistical learning has a long tradition in biodiversity research for detecting patterns in ecological data and predicting from these patterns [@guisan_predictive_2000].
It is especially useful for quantifying species-environment relationships and for predicting the occurrence of species or species richness (predictive mapping), and therefore is an indispensable instrument for biodiversity management [@fremout_mapping_2020].
Furthermore, predictive mapping can reveal ecological changes in space and time which otherwise might be hard or impossible to detect [@lovelace_geocomputation_2019a].
Therefore, it is a crucial tool for gaining detailed insights into ecosystem dynamics, which other more descriptive methods such as global biodiversity measures and ordination techniques cannot capture.
We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.

17. **2.3.2: By using NDVI as the only proxy for assemblages, all you do is to assume that diversity increases with vegetation cover and/or biomass (as the NDVI is increasing with cover/biomass but not with diversity if the biomass is the same). Probably there is a good correlation between cover and diversity but this scientific explanation should be made clear here. However, in line 256-257 you write that 2007 was wetter and biomass increased (consequently increased NDVI) but diversity decreased. How does this fit to your NDVI~diversity model? **

In fact, we have modeled the first axis of the DCA as a function of the interaction NDVI:year.
The input for the DCA was a plot-species matrix consisting of the decimal-scale transformed cover values, hence the resulting scores comprise both cover and diversity (number of species), and represent the floristic composition of the observations condensed to a single number.
Therefore, a score can also represent a plot whose number of species has decreased while the overall plant cover has increased.
In fact, Fig. 2 in the manuscript provides a visual representation how the position of one plot can change in ordination space with changing cover and species richness values.
Please refer also to the reply to the next comment in which we included a figure that shows the relationship between the response, i.e., the first DCA axis, and NDVI by year (Fig. \@ref(fig:dca-ndvi)).
Thank you again for advising us to include this figure which is crucial since it makes transparent the observed relationship between the scores of the first axis and NDVI and how this relationship changes between years.

18. **Also, why do you include the year as predictor? By doing so, you don’t set up a model that would be in any way applicable beyond the years you sampled. Why are you not assuming that the relationship between NDVI and diversity stays the same independently of the year? **

We included the year-NDVI interaction because we wanted to allow the model to have a different intercept and a different slope for each year (Fig. \@ref(fig:dca-ndvi)).
Additionally, we used a spline smoother to allow for the observed non-linearity (Fig. \@ref(fig:dca-ndvi)).
And you are absolutely right that it is crucial to give the reader the opportunity to have a look at the relationship between NDVI and the response variable.
Therefore, we have also put Fig. \@ref(fig:dca-ndvi) into the Supplementary Information and have cross-referenced it in the main text.

```{r dca-ndvi, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between the response (DCA1) and NDVI by year. To aid visual interpretation, line smoothers were added.", out.width="70%"}
data = readRDS(here("images/15_model_input_data.rds"))
ggplot(data, aes(x = ndvi, y = DCA1, group = year)) +
  geom_point(aes(color = as.factor(year)), alpha = 0.4) + 
  labs(color = "year") +
  xlab(label = "NDVI") +
  ylab(label = "DCA1") +
  geom_smooth(aes(color = as.factor(year)), se = FALSE) + 
  theme_bw() +
  coord_fixed(ratio = 0.15)
```

The aim of our study was to visualize by means of predictive mapping how the floristic composition changed between years, i.e., how different ENSO phases shape the spatial distribution of vegetation assemblages.
Therefore, we were not interested in predicting unobserved or future years.
But if we were, it would be possible to do so by taking a mean intercept and a mean slope.
However, doing so, would cancel out the majority of the interannual differences.

To make our aims clearer, we have rephrased the corresponding hypothesis as follows:

(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].

19. **More general question: why do you need the spatial prediction? Isn’t it sufficient to look at the plots that were sampled to test your hypotheses?**

Only the predictive mapping approach reveals how different ENSO episodes changed the spatial distribution of floristic composition and diversity.
An ordination by itself only reveals that diversity has changed along the gradient but cannot reveal spatial patterns.
But thank you for pointing us to formulate our hypotheses and aims in more detail.
Please refer also to the replies to comments #16, 18 and to the reply to comment #4 of reviewer #2.
And please note also that we discuss the outcome of the spatial prediction in detail in section *4.2 Floristic gradient mapping across ENSO episodes*.

20. **Numbers in Fig.2 not readable**

We now show only a few plot numbers which should illustrate how the position in ordination space has changed for a few selected examples throughout the years (Fig. \@ref(fig:dca)).
